\chapter*{Preface}
Thank you everybody
\chapter*{Summary}
Something something modeling stellar winds
\chapter*{Summary in layman's terms}
Most matter in the universe is not gaseous, liquid or solid. Most matter in the universe is in a fourth state: plasma, ionised gas. Stars, nebulae and some interstellar clouds all consist of this state of matter. Modelling the movement of plasmas is done with sophisticated computer codes, as its behaviour can be very chaotic. Another important player in the universe is radiation. Most objects we see in the skies, the sun during the day and the stars during the night are emitting strong electromagnetic fields.\\

As you may know, matter and radiation can interact. Not only can radiation heat materials (that's why it is warm when the sun is out), objects also cool when they emit themselves, and radiation can exert a force on an object: think about solar sails being used on satellites. In many astrophysical contexts this interaction is of great importance. For example: the most massive stars blow away parts of their outer layers due to this radiation force; The temperature of objects surrounding stars is set by the stellar radiation field like the temperature on earth is set by the sun.\\

This thesis focusses on adapting a computer code used to model plasmas and add a parts to it which describe the interaction between radiation and plasma. After writing the code, of course it has to be tested to check if everything works. The end product is not only a working and tested computer code, time was also spent on modelling the matter being blown away from stellar surfaces and the insides of a star.  